Article Title

### 2013-07-23 sections/Abstract.tex
converted to Frontiers' template

Linear correlations between the spiking activity of pairs of neurons are among the most familiar and useful descriptive statistics of neural activity.
Correlation matrices from multineuronal recordings contain richer information than available from an equivalent number of isolated pairs:
The basic structure of population activity can manifest in various low-dimensional representations of the covariance matrix.  The optimal choice of such reduced representation  depends profoundly on how well it can capture regularities in the underlying data generating process. 
For example, presumed diffuse correlated input onto all neurons implies a low-rank parameterization such as used in  factor analysis whereas presumed strong pairwise linear associations imply sparse partial correlations.  
Composite  `sparse+low-rank partial correlation' parameterizations allow combining common correlated input with strong pairwise associations.

To infer the optimal low-dimensional representation of neural correlations in a specitifc neural microcircuit, we compared the performance of several estimators of the covariance matrix on the activity of 100--300 neurons in mouse visual cortex: sample covariance, covariance shrinkage, factor analysis, sparse partial correlations, and sparse+low-rank partial correlations. 
The instantaneous firing rates, binned at 200 ms, were inferred from the somatic calcium signals acquired with fast 3D random-access two-photon microscopy.   
As expected, covariance shrinkage reliably outperformed the sample covariance estimate.
Factor analysis-based estimates significantly outperformed covariance shrinkage.
Yet sparsified partial correlations with or without an additional low-rank component significantly outperformed both factor analysis and shrinkage-based  estimators.
The superior performance of sparse partial correlation-based estimates suggests the relative importance of detailed network interactions over common input in this circuit.
The resulting sparse partial correlation matrix, we argue, serves as a better substrate for functional network analysis. 
In addition, the overall estimation improvement will propagate to any existing analyses that rely on the covariance matrix.

>>>>

Linear correlations between the spiking activity of pairs of neurons are among the most familiar and useful descriptive statistics of neural activity.
Multineuronal recordings provide far richer information than the equivalent number of neuron pairs considered in isolation.
For example, the full covariance matrix reveals the partial correlations between pairs of neurons or the common correlated activity involving the entire population.
Usual estimates of covariance matrix parameters require large samples for convergence.
Convergence can be improved by regularization, i.e.~by imposing some kind of structure.  
The optimal choice of the regularization approach is largely an empirical question as it depends profoundly on how well the imposed structure can still capture the dominant underlying physiogical interactions.
For example, presumed diffuse correlated input onto many neurons motivate low-rank parameterizations such as in factor analysis.  
Conversely, presumed strong pairwise linear associations between neurons motivate regularization by sparsifying the inverse covariance matrix.
Finally, `sparse+low-rank' inverse covariance parameterizations allow combining diffuse correlations and strong pairwise associations.

To infer the optimal structure of neural correlations in a specitifc neural microcircuit, we compared the performance of several covariance estimators on the activity of 100--300 neurons in mouse visual cortex: sample covariance, Ledoit-Wolf covariance shrinkage, factor analysis, sparse inverse covariance, and sparse+low-rank inverse covariance.
The instantaneous firing rates, binned at 200 ms, were inferred from the somatic calcium signals acquired with fast 3D random-access two-photon microscopy.   
Each covariance estimator was optimized and evaluated by cross-validation.
Unsurprisingly, covariance shrinkage reliably outperformed the sample covariance estimate.
In turn, factor analysis-based estimates significantly outperformed covariance shrinkage.
Yet sparse inverse covariance with or without an additional low-rank component significantly outperformed both factor analysis and shrinkage  estimators.
The superior performance of the sparse inverse covariance estimator suggests the relative importance of detailed network interactions over common diffuse input in this circuit, although these results may vary in other circuits or in other brain states.

### 2013-07-23 sections/Introduction.tex
converted to Frontiers' template

\hl{\tiny why correlations:}
Linear correlations between the spiking activity of pairs of neurons, or simply \emph{neural correlations}, are among the most familiar descriptive statistics of neural population activity \citep{Cohen:2011}.   
For example, \emph{noise correlations}, i.e.~the correlations of stimulus response variability between pairs of neurons in sensory areas, have been of particular interest thanks, in part, to their profound theoretical implications for stimulus coding \citep{Zohary:1994,Abbott:1999,Averbeck:2006,Berens:2011}.  
Lending further credence to their role as indicators of detailed and specific functional connectivity have been a series of discoveries of nontrivial relationships between neural correlations and other aspects of circuit organization such as the physical distance separating the neurons, their synaptic connectivity and stimulus tuning similarity, cortical layer specificity, cell-type specificity, progressive changes in development and in learning, changes due to sensory stimulation and global brain states, and others \citep{Kohn:2005,Smith:2008,Kohn:2009,Goard:2009,Golshani:2009,Renart:2010,Ecker:2010,Smith:2013,Denman:2013}.

\hl{\tiny ambiguity of interpretation:}
Neural correlations do not come with ready or unambiguous mechanistic interpretations.   
Theoretical and simulation studies show that neural correlations at various temporal scales arise from multiple underlying mechanisms such as direct synaptic interactions,  common or correlated inputs, shared sensory noise, chains of multiple synaptic connections, oscillations, top-down modulation, and background network activity \citep{Perkel:1967b,Shadlen:1998,Salinas:2001,Ostojic:2009}

\hl{\tiny opportunities with multineuronal data:}
The neural correlation matrix from multineuronal recordings is greater than the sum of its parts: it can be transformed into other representations  that accentuate different aspects of population activity.  
For example, the eigenvalue decomposition of the covariance matrix, also known as principal component analysis (PCA), expresses shared activity components across the entire population;  common fluctuations of population activity of the entire circuit may be accurately represented by just a few principal components but may affect all correlation coefficients. 
In contrast, the inverse of the correlation matrix reveals the partial correlations between neuron pairs, which reflect their specific linear depedence, conditioned on the activity of all the other recorded cells;
a strong interaction between a pair of neurons may be expressly represented by a single partial correlation but its effect would propagate to multiple correlations and principal components.

\hl{\tiny motivate dimensionality reduction:}
 Covariance estimates from real high-dimensional data  are both noisy and redundant.  The amount of recorded data increases only linearly with population size whereas the number of free parameters in the covariance matrix increases quadratically; as a result, empirical covariance estimates for multidimensional data are  accumulate substantial redundancy and noise.

To reduce the estimation error, statistically efficient estimators are regularized.  One regularization approach is to shrink the estimate toward a less variable low-dimensional estimate \citep{Schafer:2005}.  Another regularization approach, often called \emph{sparsification}, reduces the dimensionality of the estimate by setting its least significant parameters to zero. 
Regularized estimates perform favorably on new samples of data compared to the full model, provided the amount of regularization is chosen carefully.

\hl{\tiny choosing between regularization schemes}
Estimator improvements due to regularization depend profoundly on the parameterization to which it is applied.  Sparsification of the eigenspectrum produces a low-rank estimate of the covariance matrix also known as truncated PCA \citep{Rothman:2008} or, allowing for additional independent variances, as factor analysis (FA) \citep{Fan:2008}.  Sparsification of the inverse covariance is known as covariance selection \citep{Dempster:1972,Friedman:2008} and is related to finding a graphical model in which a large fraction of neuron pairs are conditionally independent of each other.  Representations that are capable of concentrating the significant real effects in a small number of parameters will reap greatest benefits from sparsification over ones that distribute the real effects widelyacross many parameters.  Random noise tends to be broadly distributed across all parameters in any parameterization.

\hl{\tiny our approach and summar of findings} 
In this study, we compared the performance of several covariance estimation schemes for spatially compact groups of 150--300 neurons in layers 2-4 in mouse primary visual cortex during visual stimulation.   The performance of the estimators was evaluted by computing the mean squared error between the optimized covariance estimate fitted to training data and the non-regularized covariance matrix from a separate test data set.  Low-rank covariance estimates performed significantly better than shrinkage estimators, but estimators with sparse partial correlations were more efficient still. Typically, between 3 and 16\% of neuronal paris were connected by non-zero partial interactions.  Mixed sparse estimators with sparse partial correlations and low-rank components performed comparably.

>>>>

\hl{\tiny why correlations:}
Linear correlations between the spiking activity of pairs of neurons, or simply \emph{neural correlations}, are among the most familiar descriptive statistics of neural population activity \citep{Cohen:2011}.   
For example, \emph{noise correlations}, i.e.~the correlations of stimulus response variability between pairs of neurons in sensory areas, have been of particular interest thanks, in part, to their profound theoretical implications for stimulus coding \citep{Zohary:1994,Abbott:1999,Averbeck:2006,Berens:2011}.
Lending further credence to their role as indicators of detailed and specific functional organization have been a series of discoveries of nontrivial relationships between neural correlations and other aspects of circuit organization such as the physical distance separating the neurons, their synaptic connectivity and stimulus tuning similarity, cortical layer specificity, cell-type specificity, progressive changes in development and in learning, changes due to sensory stimulation and global brain states, and others \citep{Kohn:2005,Smith:2008,Kohn:2009,Goard:2009,Golshani:2009,Renart:2010,Ecker:2010,Smith:2013,Denman:2013}.

\hl{\tiny ambiguity of interpretation:}
Neural correlations do not come with ready or unambiguous mechanistic interpretations.   
Theoretical and simulation studies show that neural correlations at various temporal scales may arise from any combination of underlying mechanisms such as direct synaptic interactions,  common or correlated inputs, shared sensory noise, chains of multiple synaptic connections, oscillations, top-down modulation, and background network activity \citep{Perkel:1967b,Shadlen:1998,Salinas:2001,Ostojic:2009,Rosenbaum:2011}

\hl{\tiny opportunities with multineuronal data:}
Early studies of neural correlations were based on measurements from isolated pairs of neurons and their impact on coding was extrapolated to entire populations \citep{Shadlen:1998,Zohary:1994}.  Modern massively multineuronal recordings allow estimations of entire correlation matrices from large neuronal populations, which provide more information than the equivalent number of pairwise correlations assessed in isolation.
After all, the correlation matrix is greater than the sum of its parts: it can be transformed into other representations  that accentuate different aspects of its structure and may suggest different mechanistic interpretations.  
For example, the eigenvalue decomposition, also known as the principal component analysis, of the covariance matrix expresses shared correlated activity components across the entire population;  common fluctuations of population activity of the entire circuit may be accurately represented by just a few principal components but will affect all correlation coefficients. 
In contrast, the off-diagonal elements of the inverse of the correlation matrix constitute scaled partial correlations between neuron pairs, which reflect their specific linear dependencies, after accounting for all other recorded cells;
a strong interaction between a pair of neurons may be expressly represented by a single partial correlation but its effects propagate to multiple correlations and principal components.

\hl{\tiny motivate regularization} Estimation of parameters of the covariance matrix from data is inherently challenging, increasingly for very high-dimensional data. As the amount of data grows only linearly with population size, the number of free parameters in a complete representation of the covariance matrix increases quadratically, multiplying opportunities for spurious patterns to emerge among the correlations. 
Although the estimation error of each covariance coefficient, in isolation, is unaffected by the dimensionality of the data, the estimation error of other parameters of the covariance matrix increases with population size.  
For example, the large eigenvalues of in the sample covariance matrix are biased upward while small eigenvalues are biased downward, with both variance and bias increasingly large with increasing data dimensionality \citep{Hayes:1981}.  Similarly, the coefficients of the inverse covariance matrix require  larger sample sizes to be estimated accurately in high-dimensional data.

\hl{\tiny forms of regularization} The convergence of the estimate of multivariate parameters to their true values can be improved through \emph{regularization},  Regularization is the deliberate biasing or \emph{shrinkage} of the empirical estimate toward a less variable \emph{target estimate} \citep{Schafer:2005} or the deliberate restriction of the estimate to a lower-dimensional space \citep{Dempster:1972,Fan:2008,Friedman:2008,Rothman:2008}. 
\hl{\tiny The two definitions overlap and are often equivalent, but not always, according to James, but I would like to see an example of shrinkage that is not, effectively, a low dimensional parameterization of the original problem.} 
One intuitive explanation of regularization is as the addition of some  prior knowledge about the likely forms of the covariance matrix. Curiously, though, regularization need not reflect any accurate knowledge about the nature of covariance matrices in general or in the particular domain. Some improvement arises due to \emph{Stein's phenomenon}, i.e. the favorable tradeoff between bias and variability \emph{anytime} a moderate bias toward a less variable target is introduced.  
However, the incorporation into the regularization scheme of real knowledge about the probable forms of covariance matrices for a specific situation will likely confer additional advantage and improve the convergence more substantially than an \emph{ad hoc} regularizer.

%Estimator improvements due to regularization depend profoundly on the parameterization to which it is applied.  Sparsification of the eigenspectrum produces a low-rank estimate of the covariance matrix also known as truncated PCA \citep{Rothman:2008} or, allowing for additional independent variances, as factor analysis (FA) \citep{Fan:2008}.  Sparsification of the inverse covariance is known as covariance selection \citep{Dempster:1972,Friedman:2008} and is related to finding a graphical model in which a large fraction of neuron pairs are conditionally independent of each other.  Representations that are capable of concentrating the significant real effects in a small number of parameters will reap greatest benefits from sparsification over ones that distribute the real effects widelyacross many parameters.  Random noise tends to be broadly distributed across all parameters in any parameterization.

Two questions must be answered in the choice of the regularization scheme: what kind of regularization and how much regularization. The answer to the first question is domain-specific and is often motivated by theories of the origins of correlations in the specific subject domain.

\hl{\tiny approach and summary of findings} 
In this study, we compared the performance of several covariance estimation schemes for spatially compact groups of 150--300 neurons in layers 2-4 in mouse primary visual cortex during visual stimulation.   The performance of the estimators was evaluted by computing the mean squared error between the optimized covariance estimate fitted to training data and the non-regularized covariance matrix from a separate test data set.  Low-rank covariance estimates performed significantly better than shrinkage estimators, but estimators with sparse partial correlations were more efficient still. Typically, between 3 and 16\% of neuronal paris were connected by non-zero partial interactions.  Mixed sparse estimators with sparse partial correlations and low-rank components performed comparably.

### 2013-07-23 sections/Theory.tex
converted to Frontiers' template

\subsection*{Problem setup}
Let $x_1,\ldots,x_n \in \mathbb R^{p\times 1}$ denote a sample of $n$ observations of instantaneous activity (e.g.~spiking) rates of $p$ neurons averaged in time bins $i=1,\ldots,n$.

For simplicity, we assume that observations are sampled from a shared true distribution $x_1,\ldots,x_n \sim f$. 
\hl{\tiny explain: the derivations remain valid even without such a strong assumption.  We account for overdispersion due to temporal dependence and shuffle chunks of data to account for nonstationarities.}

@@

In particular, we consider the mean-squared loss
\begin{equation}
\mathcal L_e( S, \Sigma) =  \| S - \Sigma\|^2_F 
\end{equation}
where $\| \cdot \|_F$ denotes the Frobenius norm;

and the multivariate normal log likelihood loss
\begin{equation}
\mathcal L_g(S,\Sigma) = \ln |S| +  \Tr(S^{-1} \Sigma)
\end{equation}
where $|S|$ denotes the determinant of matrix $S$.

>>>>

\subsection{Problem setup}
Let $x_1,\ldots,x_n \in \mathbb R^{p\times 1}$ denote a sample of $n$ observations of instantaneous activity (e.g.~spiking) rates of $p$ neurons averaged in time bins $i=1,\ldots,n$.

For simplicity, we assume that observations are sampled from a shared true distribution $x_1,\ldots,x_n \sim f$. 
\hl{\tiny explain: the derivations remain valid even without such a strong assumption.}

@@

In particular, we consider the sum of squared-error loss
\begin{equation}
\mathcal L_e( S, \Sigma) =  \| S - \Sigma\|^2_F 
\end{equation}
where $\| \cdot \|_F$ denotes the Frobenius norm;

Another useful loss denoted here as $\mathcal L_g$ is based  on the log likelihood of the multivarate normal distribution 
$$\ln L(\Sigma \cond S) = - \tfrac 1 2 k\ln(2\pi) - \tfrac 1 2 \ln |S| - \tfrac 1 2 \Tr(S^{-1}\Sigma)$$ 
where $|S|$ denotes the determinant of matrix $S$.

The loss function is shifted so that $\mathcal L_g(\Sigma,\Sigma) = 0$:
\begin{equation}
\begin{split}
\mathcal L_g(S,\Sigma) =&  -2 \left[\ln L(\Sigma\cond S) - \ln L(\Sigma\cond \Sigma)\right]
\\=& \Tr(S^{-1}\Sigma) -\ln |S^{-1}\Sigma| - p
\end{split}
\end{equation}

Stein's loss can also be interpreted as the Kullback-Leibler divergence between the multivarate normal with covariance $\Sigma$ and another with $S$:
\begin{equation}
\mathcal L_{s} = 2 D_{KL} (\mathcal N(0,\Sigma) \| S)
\end{equation}
